\documentclass[
11pt,
fleqn,
titlepage]{article}

% Packages

\usepackage[sfdefault]{roboto}
\usepackage{authblk}
\usepackage[a4paper, headheight=14pt]{geometry}
\usepackage[document]{ragged2e}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage[section]{placeins}
\usepackage{siunitx}
\usepackage{textgreek}
\usepackage[version=4]{mhchem}
\usepackage[babel, tracking, verbose]{microtype}
\usepackage[defaultlines=2, all]{nowidow}
\usepackage{subcaption}
\captionsetup{
    labelfont=bf,
    singlelinecheck=false,
    justification=RaggedRight}
\captionsetup[subfigure]{
    labelformat=brace}
\renewcommand\thesubfigure{\roman{subfigure}}
\usepackage{polyglossia}
\setdefaultlanguage[variant=american]{english}
\usepackage{csquotes}
\usepackage[style=apa]{biblatex}
\addbibresource{bibliography.bib}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{soul}
\usepackage[labelfont=bf]{caption}
\usepackage{tabularx}
\usepackage{threeparttablex}
\usepackage[imakeidx]{xindex}
\makeindex[intoc]
\usepackage[pdfa]{hyperref}
\usepackage{colorprofiles}
\hypersetup{
    pdfinfo={
        Title={Genetic analysis of phenotypic variation measured in the Oregon Wolfe Barley collection},
        Author={Halide Aydin, Aliz Fodor, Ingo Giebel, Alessia Russo},
        Subject={Research project QBio305: Population and Quantitative Genetics},
        Keywords={Heinrich-Heine-Universit채t D체sseldorf, WS 2023/2024, QBio305: Population and Quantitative Genetics, Research project, Genetic analysis of phenotypic variation, Oregon Wolfe Barley collection, Instructors: Prof. Dr. Maria von Korff Schmising, Dr. Timo Hellwig, Dr. Agatha Walla}},
        colorlinks,
        linkcolor=blue,
        linktoc=all}
\usepackage[ocgcolorlinks, tikz]{ocgx2}
\usepackage{biocon}
\newplant{Hb}{genus=Hordeum, epithet=bulbosum}
\sloppy

% Header

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{Oregon Wolfe Barley collection: genetic analysis of phenotypic variation}
\fancyhead[R]{\thepage}
\fancypagestyle{plain}{}

% Title & author

\title{Genetic analysis of phenotypic variation measured in the Oregon Wolfe Barley collection}
\author{Halide Aydin}
\author{Aliz Fodor}
\author{Ingo Giebel}
\author{Alessia Russo}
\affil{QBio305: Population and Quantitative Genetics}
\affil{Institute of Plant Genetics, Heinrich-Heine-Universit채t D체sseldorf}
\affil{Prof. Dr. Maria von Korff Schmising, Dr. Timo Hellwig, Dr. Agatha Walla}
\date{December 22, 2023}

\begin{document}
    \maketitle

    \tableofcontents

    \clearpage
    \renewcommand*\listtablename{List of tables}
    \listoftables
    \renewcommand*\listfigurename{List of figures}
    \listoffigures

    \clearpage

    \input{introduction}

    \input{materials_and_methods}

    \input{results_and_discussion}

    \clearpage
    \printbibliography[heading=bibintoc]

    \clearpage
    \appendix
    \input{appendix_a}
    \input{appendix_b}
    \input{appendix_c}

    \clearpage
    \printindex
\end{document}